---
title: "Smoothing vs. Outlier Removal in AXA's Driver Telematics Analysis competition 
(or how to fix the Hyperspace Jumps)"
author: "Mario Vivero"
date: "1/10/2015"
output: html_document
---
  Smoothing vs. Outlier Removal 
========================================================
This article is meant as a graphical comparison between smoothing techniques and simple outlier detection. 
Most popular data derived from GPS include[1]: 
-Average velocity
-Average running velocity except stop
-Positive acceleration kinetic energy change per unit mass per unit
-Average acceleration
-Average deceleration
-Maximum velocity
-Maximum acceleration
etc.
For most of these features a proper cleaning must be done due to measurement errors or in the case of the AXA competition
several data points were removed due to privacy concerns. The most notorious example is Driver 1 Trip 136
as was discussed on the Kaggle forums:
http://www.kaggle.com/c/axa-driver-telematics-analysis/forums/t/11288/hyperspace-jumps-or-paused-gps-tracking

Driver 1 Trip 136 , around sec 275
-980.2,250.9
-1000.3,259.8
-1545.0,397.5 
-1569.0,382.1

This will be the trip that I will use to make the graphical comparison between techniques.

Libraries, Functions and Data
------------

```{r results='hide', message=FALSE}
#Libraries, directories, options and extra functions----------------------
require("data.table")
require("parallel")
require("h2o")
require("ggplot2")
require("prospectr")

#Set Working Directory
driversDirectory <- "/home/wacax/Wacax/Kaggle/AXA Driver Telematics Analysis/Data/drivers" #change this to your own directories

#Detect available cores
numCores <- detectCores()

```

We will start by visualizing all the 200 trajectories of a single driver.
```{r fig.width=16, fig.height=16}
#Visualization of Trajectories for driver number 1; mcapply only works on linux
#for more information about parallel alternatives on windows see: http://blog.dominodatalab.com/simple-parallelization/
results <- mclapply(seq(1, 200), function(file, driverID){
  tripCoordinates <- fread(file.path(driversDirectory, driverID, paste0(file, ".csv")))
  return(cbind(tripCoordinates, rep(file, nrow(tripCoordinates))))
}, mc.cores = numCores, driverID = 1)
print(paste0("Driver number: ", 1, " processed"))
dataTableReady2Plot <- do.call(rbind, results)
#All hail the FSM
qplot(x, y, data = dataTableReady2Plot, colour = V2, geom = "point")
```

As can be seen from the trajectories, there are two trajectories with missing fragments. So for trajectories like these we will compare techniques that can yield appropiate data.

Trajectory of Driver 1 Trip 136 
------------
```{r fig.width=16, fig.height=16}
driverViz <- 1
fileViz <- 136
tripCoordinates <- fread(file.path(driversDirectory, driverViz, paste0(fileViz, ".csv")))
ggplot(as.data.frame(tripCoordinates), aes(x = x, y = y)) + geom_point()
print(paste0("Driver number: ", driverViz, " trip number ", fileViz, " processed"))
```
To determine the speed between each trajectory point we then use the following formula.
The output of this formula will be a vector of velocities in km/h of length corresponding to the number of points in each trajectory minus 1

```{r fig.width=16, fig.height=16}
speed2Plot <- 3.6 * sqrt(diff(tripCoordinates$x)^2 + diff(tripCoordinates$y)^2)
head(speed2Plot, 50)
```
We now explore the distribution of velocities in trip 136 of driver # 1

```{r fig.width=16, fig.height=16}
ggplot(as.data.frame(speed2Plot), aes(x = speed2Plot)) + geom_density()
```
The graph above doesn't look correct. There is a single value corresponding to aproximately 2500 km/h that makes this distribution look funny. So now we will compare different techniques to make this distribution look better.
A common technique is to use a rolling mean or a rolling median.

```{r fig.width=16, fig.height=16}
driverViz <- 1
fileViz <- 136
tripCoordinates <- fread(file.path(driversDirectory, driverViz, paste0(fileViz, ".csv")))
speed2Plot <- 3.6 * sqrt(diff(tripCoordinates$x, 20, 1)^2 + diff(tripCoordinates$y, 20, 1)^2) / 20
ggplot(as.data.frame(speed2Plot), aes(x = speed2Plot)) + geom_density()

```

Unfortunately, using a rolling mean or a median will produce a tail of unexistent data, in our case the neighbors of the point close to the jump in the trajectory. A different window could be used to optimize the velocities but it will be almost imposible to find an appropiate window for all drivers in this dataset.

Let's try now fittin a polinomial using the Savitzky-Golay filter. More information about the Savitzky-Golay filter can be found on http://rpubs.com/wacax/33342 and [3]Beebe at. al., Chemometrics - A Practical Guide

```{r fig.width=16, fig.height=16}
driverViz <- 1
fileViz <- 136
tripCoordinates <- fread(file.path(driversDirectory, driverViz, paste0(fileViz, ".csv")))
speed2Plot <- as.data.frame(savitzkyGolay(3.6 * sqrt(diff(tripCoordinates$x)^2 + diff(tripCoordinates$y)^2), p = 3, w = 11, m = 0))
names(speed2Plot) <- "speed2Plot"
ggplot(as.data.frame(speed2Plot), aes(x = speed2Plot)) + geom_density()
```

Again, not a very good idea since it yields some negative values and high positive values.
Let's now compare a 5 sigma outlier removal technique.

```{r fig.width=16, fig.height=16}
#Remove data above 5 sigmas (5 standard deviations)
driverViz <- 1
fileViz <- 136
tripCoordinates <- fread(file.path(driversDirectory, driverViz, paste0(fileViz, ".csv")))
speed2Plot <- 3.6 * sqrt(diff(tripCoordinates$x)^2 + diff(tripCoordinates$y)^2)
#Remove data above 5 sigmas (5 standard deviations)
speed2Plot2 <- speed2Plot[!speed2Plot > mean(speed2Plot) + sd(speed2Plot) * 5]
ggplot(as.data.frame(speed2Plot2), aes(x = speed2Plot2)) + geom_density()
```

Now this distribution plot shows the exact velocities without strange values which correspond to the data removed from trajectories.
Let's now compare this preprocessed trajectory with a trajectory without strange values.

```{r fig.width=16, fig.height=16}
driverViz <- 1
fileViz <- 1
tripCoordinates <- fread(file.path(driversDirectory, driverViz, paste0(fileViz, ".csv")))
speed2Plot <- 3.6 * sqrt(diff(tripCoordinates$x)^2 + diff(tripCoordinates$y)^2)
ggplot(as.data.frame(speed2Plot), aes(x = speed2Plot)) + geom_density()

```
```{r fig.width=16, fig.height=16}
#Remove data above 5 sigmas (5 standard deviations)
speed2Plot2 <- speed2Plot[!speed2Plot > mean(speed2Plot) + sd(speed2Plot) * 5]
ggplot(as.data.frame(speed2Plot2), aes(x = speed2Plot2)) + geom_density()
```

It's evident now that outlier removal yields the most similar results to the trajectories without strange data.
The outlier removal technique can be used for other features as well such as accelerations, angles and others.


Bibliography
------------
[1]Wang R., Lukic S. M; Review of Driving Conditions Prediction and Driving Style Recognition Based Control Algorithms for Hybrid Electric Vehicles, 978-1-61284-246-9/11/ ©2011 IEEE http://178.22.59.152/documents/Driving%20Style%20Recognition%20Based.pdf
[2]http://blog.dominodatalab.com/simple-parallelization/
[3]Beebe at. al., Chemometrics - A Practical Guide, March 1998, ISBN: 978-0-471-12451-1
